Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)

### 2017-02-15 csf-2017/paper.tex
removed acknowledgement and authors
Rephrase control flow integrity to control flow correctness.

\author{%
\IEEEauthorblockN{Lau~Skorstengaard}
\IEEEauthorblockA{Aarhus University, Denmark\\
Email: lau@cs.au.dk} \and
\IEEEauthorblockN{Dominique~Devriese}
  \IEEEauthorblockA{iMinds-DistriNet, KU Leuven, Belgium\\
Email: dominique.devriese@cs.kuleuven.be} \and
\IEEEauthorblockN{Lars~Birkedal}\IEEEauthorblockA{Aarhus University, Denmark\\
Email: birkedal@cs.au.dk}}

\maketitle

\begin{abstract}  
  Capability machines provide security guarantees at machine level
  which makes them an interesting target for compilation schemes that aim to
  provably enforce control-flow integrity, encapsulation of local
  state, and such properties.
  We provide a formalisation of a representative capability machine
  and propose a novel stack management discipline for enforcing
  control-flow integrity and encapsulation of local state on it. To prove
  these properties, we provide a logical relation that semantically
  captures capability safety. The capability safety defined by the
  logical relation goes beyond the
  proposed stack management discipline and can be used to reason about
  any capability safe program.
\end{abstract}

@@

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control flow
integrity, local state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control flow integrity and
local state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) or rely on techniques like
machine code rewriting
(e.g.,~\cite{wahbe_efficient_1993,abadi_control-flow_2005}), machine code
verification (e.g.,~\cite{morrisett_system_1999}), virtual machines with a native
stack (e.g.,~the JVM~\citep{lindholm_java_2014}) or
randomization~\citep{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

@@

In contrast to the situation for commodity processors, the security
primitives on capability machines lend themselves well to the
enforcement of properties like local state encapsulation and
control-flow integrity. Potentially, they will enable new compilation
schemes that enforce such properties in a way that is efficient but
also 100\% watertight (ideally evidenced by a mathematical proof,
guaranteeing that we do not end up in a new attack-defense arms
race). However, there is still a lot that needs to happen to attain
this goal. For example, it is quite non-trivial to devise a
compilation scheme adapted to the details of a specific language's
notion of encapsulation (e.g., encapsulation of private member
variables in OO languages often behaves quite differently than
encapsulation of private state in ML-like languages). And even if such
a scheme were defined, a formal proof depends on a formalisation of
the encapsulation provided by the capability machine at hand.

A similar problem is the enforcement of control-flow integrity on capability
machines. An interesting approach is taken in
CheriBSD~\citep{Watson2015Cheri}: It drops the standard
contiguous C stack and splits it into a central, trusted shadow stack, managed by
trusted call and return instructions, and disjoint, per-component
stacks. To prevent illegal use of stack references,
the approach relies on \emph{local capabilities}, a type of capabilities offered
by CHERI that can be passed over a function invocation \emph{temporarily}, for
the duration of the invocation, and revoked afterwards. However, a lot of
details remain to be investigated (e.g., how to combine a trusted shadow stack with
(potentially) cross-domain function pointers in a scalable way?) and there is no
argument that this enforcement of control-flow integrity is watertight.

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow integrity on capability
machines.  Specifically, we make the following contributions:
\begin{itemize}
\item In Section~\ref{sec:capab-mach-with}, we formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  integrity and encapsulation of local state on the stack in Section~\ref{sec:stack-and-return-pointer}. The discipline relies
  solely on local capabilities (used to represent stack and return pointers) and
  does not require OS support in the form of a trusted shadow stack or special
  call and return instructions. It also does not require separate per-component
  stacks and instead works with a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., (potentially)
  cross-component function pointers) and can be efficient assuming
  only
  one additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed
  logical relation with recursively defined Kripke worlds. It is
  inspired by one used in \cite{Devriese:2016ObjCap} for
  capability safety in a JavaScript-like lambda calculus, but has two
  novel technical characteristics. The natural CPS style of assembly
  code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal
  closure that has been previously used for lambda calculi. Perhaps
  more surprisingly, we show that the special nature of local
  capabilities can be elegantly expressed using a variant of an idea
  that has been used by Dreyer et. al.~\cite{Dreyer:jfp12} to express
  well-bracketedness guarantees in a lambda calculus: Kripke worlds
  with both a public and a private future world relation. Interestingly,
  the two future world relations generally work very similarly to
  Dreyer et al.'s, except that they turn up in quite a different place
  in the logical relation. The logical relation and theorem is presented in Section~\ref{sec:logical-relation}.
\item We demonstrate the power of our results in Section~\ref{sec:examples} by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow integrity guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability safety and our stack management discipline.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

\section{Stack and return pointer management using local capabilities}
\label{sec:stack-and-return-pointer}
%\lau{Do we want to use the term ``pointer'' at any point? I have also used it, but I guess it might create some confusing to whether we have traditional pointer as well as capabilities?}
One of the contributions in this paper is a demonstration that local
capabilities on a capability machine can be used in a stack and return pointer
management scheme that enforces control flow integrity in a way that is provably
watertight, potentially efficient, does not rely on a trusted central stack
manager, and with support for exposing higher-order interfaces to an adversary.
In this section, we explain this discipline. We start by explaining the
high-level approach, the different issues that need to be taken into account,
starting with first-order interfaces and moving onto a higher-order setting
next. After that, we define a number of reusable macro-instructions that
can be used to conveniently apply the proposed discipline in subsequent
examples.

@@

\section{Examples}
\label{sec:examples}
In this section, we demonstrate how our formal model of capability
safety allows us to prove local state encapsulation and control-flow
integrity properties for challenging program examples. The security
measures of Section~\ref{sec:stack-and-return-pointer} are deployed to
ensure these properties. Since we are dealing with assembly language,
there are many details to the formal treatment, and therefore we
necessarily have to omit some details in the lemmas we state here.
The interested reader can find all the technical details in the
technical appendix~\cite{technical_appendix}.

@@

\subsection{Stack and return pointer discipline}
\textbf{Formulating control flow integrity} The attentive reader may notice
that, while we claim that our stack and return pointer discipline enforces
control flow integrity, we do not prove a general theorem to show that this is
the case. The reason for this is that it is not clear what such a theorem should
look like. Formulations in terms of a control flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on.\footnote{It is a
  bit unfortunate that the general term ``control flow integrity'' is used for a
  property that only captures partial correctness of control flow.} In fact, our
examples show that our logical relation imply a form of control flow integrity
that is stronger than such formulations, although this is not made very
explicit. As future work, we are considering to look at a more explicit and
useful way to formalise well-bracketed control flow. The idea would be to define
a variant of our capability machine with built-in call and return instructions
and well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our stack and return pointer
discipline is fully abstract~\cite{abadi_protection_1998}, which roughly means
that the lower-level machine does not allow interactions with the compiled code
that do not correspond to an interaction with the original code in the
high-level machine.

@@

More closely related to our work are papers that use a trusted stack manager and
some form of memory isolation to enforce control flow correctness as part of a
secure compilation
result~\citep{patrignani_modular_2016-1,juglaret_beyond_2016-1}. Our work
differs from theirs in that we use a different form of low-level security
primitive (a capability machine with local capabilities rather than a machine
with a primitive notion of compartments) and we do not use a trusted stack
manager component, but a decentralised calling convention based on local
capabilities. Also, both prove a secure compilation result from a high-level
language, which clearly implies a general form of control flow correctness,
while we define a sound logical relation that can be used to reason about
specific programs that rely on well-bracketed control flow.

@@

El-Korashy also defined a formal model of a capability machine, namely CHERI,
and uses it to prove a compartmentalization
result~\citep{akram_el-korashy_formal_2016} (not implying control flow
correctness). He also proves soundness of control-flow integrity (see above)
adapted to the capability machine, a result that seems orthogonal to the fact
that the machine uses capabilities.

\section{Conclusion}
\label{sec:conclusion}
\todo[inline]{Insert the correct URL for the technical report in the bibtex file!}
\section*{Acknowledgements}
\label{sec:acknowledgements}

This research was supported in part by the ModuRes Sapere Aude Advanced Grant from The Danish Council for Independent Research for the Natural Sciences (FNU).
Dominique Devriese holds a postdoctoral fellowship from the Research Foundation - Flanders (FWO).

%\bibliographystyle{IEEEtran}
\bibliographystyle{plainnat}
\bibliography{IEEEabrv,references}

>>>>

% \author{%
% \IEEEauthorblockN{Lau~Skorstengaard}
% \IEEEauthorblockA{Aarhus University, Denmark\\
% Email: lau@cs.au.dk} \and
% \IEEEauthorblockN{Dominique~Devriese}
%   \IEEEauthorblockA{iMinds-DistriNet, KU Leuven, Belgium\\
% Email: dominique.devriese@cs.kuleuven.be} \and
% \IEEEauthorblockN{Lars~Birkedal}\IEEEauthorblockA{Aarhus University, Denmark\\
% Email: birkedal@cs.au.dk}}
\author{}
\maketitle

\begin{abstract}  
  Capability machines provide security guarantees at machine level
  which makes them an interesting target for compilation schemes that aim to
  provably enforce control-flow correctness, encapsulation of local
  state, and such properties.
  We provide a formalisation of a representative capability machine
  and propose a novel stack management discipline for enforcing
  control-flow correctness and encapsulation of local state on it. To prove
  these properties, we provide a logical relation that semantically
  captures capability safety. The capability safety defined by the
  logical relation goes beyond the
  proposed stack management discipline and can be used to reason about
  any capability safe program.
\end{abstract}

@@

Compromising software security is often based on attacks that break programming
language properties relied upon by software authors, such as control-flow
correctness, local state encapsulation, etc. Commodity processors offer little
support for defending against such attacks: they offer security primitives with
only coarse-grained memory protection and limited compartmentalization
scalability. As a result, defenses against attacks on control-flow correctness
and local state encapsulation are either limited to only certain common forms of
attacks (leading to an attack-defense arms race) or rely on techniques like
machine code rewriting
(e.g.,~\cite{wahbe_efficient_1993,abadi_control-flow_2005}), machine code
verification (e.g.,~\cite{morrisett_system_1999}), virtual machines with a
native stack (e.g.,~the JVM~\citep{lindholm_java_2014}) or
randomization~\citep{forrest_building_1997}. The latter techniques essentially
emulate protection techniques on existing hardware, at the cost of performance,
system complexity and/or security.

@@

In contrast to the situation for commodity processors, the security
primitives on capability machines lend themselves well to the
enforcement of properties like local state encapsulation and
control-flow correctness. Potentially, they will enable new compilation
schemes that enforce such properties in a way that is efficient but
also 100\% watertight (ideally evidenced by a mathematical proof,
guaranteeing that we do not end up in a new attack-defense arms
race). However, there is still a lot that needs to happen to attain
this goal. For example, it is quite non-trivial to devise a
compilation scheme adapted to the details of a specific language's
notion of encapsulation (e.g., encapsulation of private member
variables in OO languages often behaves quite differently than
encapsulation of private state in ML-like languages). And even if such
a scheme were defined, a formal proof depends on a formalisation of
the encapsulation provided by the capability machine at hand.

A similar problem is the enforcement of control-flow correctness on capability
machines. An interesting approach is taken in CheriBSD~\citep{Watson2015Cheri}:
It drops the standard contiguous C stack and splits it into a central, trusted
shadow stack, managed by trusted call and return instructions, and disjoint,
per-component stacks. To prevent illegal use of stack references, the approach
relies on \emph{local capabilities}, a type of capabilities offered by CHERI
that can be passed over a function invocation \emph{temporarily}, for the
duration of the invocation, and revoked afterwards. However, a lot of details
remain to be investigated (e.g., how to combine a trusted shadow stack with
(potentially) cross-domain function pointers in a scalable way?) and there is no
argument that this enforcement of control-flow correctness is watertight.

In this paper, we make progress towards the goal of efficient and provably
watertight enforcement of encapsulation and control-flow correctness on
capability machines. Specifically, we make the following contributions:
\begin{itemize}
\item In Section~\ref{sec:capab-mach-with}, we formalise a simple but representative capability machine featuring
  local capabilities much like CHERI's, and define its operational semantics.
\item We define a novel stack management discipline for enforcing control-flow
  correctness and encapsulation of local state on the stack in
  Section~\ref{sec:stack-and-return-pointer}. The discipline relies solely on
  local capabilities (used to represent stack and return pointers) and does not
  require OS support in the form of a trusted shadow stack or special call and
  return instructions. It also does not require separate per-component stacks
  and instead works with a single, fairly standard, contiguous stack. It
  supports higher-order cross-component calls (e.g., (potentially)
  cross-component function pointers) and can be efficient assuming only one
  additional piece of processor support: an instruction for efficiently clearing
  a range of memory.
\item We provide a theorem that is (by far) the most general
  and powerful formulation of the guarantees offered by a capability machine.
  The theorem also expresses the specific guarantees offered by the machine for
  local capabilities. It is very general and not tied to a specific ABI or a
  specific way of using the system's capabilities.
\item The above theorem is formulated in terms of a unary step-indexed
  logical relation with recursively defined Kripke worlds. It is
  inspired by one used in \cite{Devriese:2016ObjCap} for
  capability safety in a JavaScript-like lambda calculus, but has two
  novel technical characteristics. The natural CPS style of assembly
  code with return pointers as continuations is modeled using a
  \emph{single} orthogonal closure, rather than the biorthogonal
  closure that has been previously used for lambda calculi. Perhaps
  more surprisingly, we show that the special nature of local
  capabilities can be elegantly expressed using a variant of an idea
  that has been used by Dreyer et. al.~\cite{Dreyer:jfp12} to express
  well-bracketedness guarantees in a lambda calculus: Kripke worlds
  with both a public and a private future world relation. Interestingly,
  the two future world relations generally work very similarly to
  Dreyer et al.'s, except that they turn up in quite a different place
  in the logical relation. The logical relation and theorem is presented in Section~\ref{sec:logical-relation}.
\item We demonstrate the power of our results in Section~\ref{sec:examples} by applying them to challenging
  examples, specifically constructed to demonstrate local state encapsulation
  and control-flow correctness guarantees in the presence of cross-component
  function pointers. The examples demonstrate both the power of our formulation
  of capability safety and our stack management discipline.
\end{itemize}
%\lau{I have left out the discussion/future work, related work and conclussion sections from the outline given with the contribution. Do you feel like there should be a dedicated outline instead and if not should we mention the last sections (I don't feel it contributes anything to mention them).}

@@

\section{Stack and return pointer management using local capabilities}
\label{sec:stack-and-return-pointer}
%\lau{Do we want to use the term ``pointer'' at any point? I have also used it, but I guess it might create some confusing to whether we have traditional pointer as well as capabilities?}
One of the contributions in this paper is a demonstration that local
capabilities on a capability machine can be used in a stack and return pointer
management scheme that enforces control-flow correctness in a way that is provably
watertight, potentially efficient, does not rely on a trusted central stack
manager, and with support for exposing higher-order interfaces to an adversary.
In this section, we explain this discipline. We start by explaining the
high-level approach, the different issues that need to be taken into account,
starting with first-order interfaces and moving onto a higher-order setting
next. After that, we define a number of reusable macro-instructions that
can be used to conveniently apply the proposed discipline in subsequent
examples.

@@

\section{Examples}
\label{sec:examples}
In this section, we demonstrate how our formal model of capability
safety allows us to prove local state encapsulation and control-flow
correctness properties for challenging program examples. The security
measures of Section~\ref{sec:stack-and-return-pointer} are deployed to
ensure these properties. Since we are dealing with assembly language,
there are many details to the formal treatment, and therefore we
necessarily have to omit some details in the lemmas we state here.
The interested reader can find all the technical details in the
technical appendix~\cite{technical_appendix}.

@@

\subsection{Stack and return pointer discipline}
\textbf{Formulating control flow correctness} The attentive reader may notice
that, while we claim that our stack and return pointer discipline enforces
control-flow correctness, we do not prove a general theorem to show that this is
the case. The reason for this is that it is not clear what such a theorem should
look like. Formulations in terms of a control-flow graph, like the one by
\cite{abadi_control-flow_2005}, do not take into account temporal properties,
like the well-bracketedness that Example~\texttt{\footnotesize{g1}} relies on. In fact, our
examples show that our logical relation imply a form of control-flow correctness
that is stronger than such formulations, although this is not made very
explicit. As future work, we are considering to look at a more explicit and
useful way to formalise well-bracketed control flow. The idea would be to define
a variant of our capability machine with built-in call and return instructions
and well-bracketed control flow built-in to the operational semantics, and then
prove that compiling such code to our machine using our stack and return pointer
discipline is fully abstract~\cite{abadi_protection_1998}, which roughly means
that the lower-level machine does not allow interactions with the compiled code
that do not correspond to an interaction with the original code in the
high-level machine.

@@

More closely related to our work are papers that use a trusted stack manager and
some form of memory isolation to enforce control-flow correctness as part of a
secure compilation
result~\citep{patrignani_modular_2016-1,juglaret_beyond_2016-1}. Our work
differs from theirs in that we use a different form of low-level security
primitive (a capability machine with local capabilities rather than a machine
with a primitive notion of compartments) and we do not use a trusted stack
manager component, but a decentralised calling convention based on local
capabilities. Also, both prove a secure compilation result from a high-level
language, which clearly implies a general form of control-flow correctness,
while we define a sound logical relation that can be used to reason about
specific programs that rely on well-bracketed control flow.

@@

El-Korashy also defined a formal model of a capability machine, namely CHERI,
and uses it to prove a compartmentalization
result~\citep{akram_el-korashy_formal_2016} (not implying control-flow
correctness). He also proves soundness of control flow integrity (see above)
adapted to the capability machine, a result that seems orthogonal to the fact
that the machine uses capabilities.

%\section{Conclusion}
\label{sec:conclusion}
\todo[inline]{Insert the correct URL for the technical report in the bibtex file!}
\begin{comment}
\section*{Acknowledgements}
\label{sec:acknowledgements}

This research was supported in part by the ModuRes Sapere Aude Advanced Grant from The Danish Council for Independent Research for the Natural Sciences (FNU).
Dominique Devriese holds a postdoctoral fellowship from the Research Foundation - Flanders (FWO).
\end{comment}
%\bibliographystyle{IEEEtran}
\bibliographystyle{plainnat}
\bibliography{IEEEabrv,references}
